---
title: "Occurrence of human infection with Salmonella Typhi in sub-Saharan Africa"
output: github_document
date: "2023-07-24"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE) 
```


### Source some utility functions first 
```{r}
library(readxl)
# source("util/clean_country_names.R")
source("util/util.R")
library(tidyverse)
read_csv("data/occ_data_20221025.csv") |>
  select(ADM0=COUNTRY, ) 

d <- data.table::fread("data/occ_data_20221025.csv")

names(d) <- c("ADM_0", "STUDY_ID", "DIAGNOSTGIC_ID", "STUDY_TYPE",
              "DIAGNOSTIC_METHODS", "STUDY_INFO")

study_types <- as.data.frame(unique(d$STUDY_TYPE))
diagnostic_methods <- as.data.frame(unique(d$DIAGNOSTIC_METHODS))

library(data.table)
# fwrite(study_types, "out/study_types.csv")
# fwrite(diagnostic_methods, "out/diagnostic_methods.csv")
d$ADM_0 <- clean_country_names(d$ADM_0)
```

### Simple descriptive statistics first 
Combine data sets into a single file
```{r}
library(readxl)
path_country_data <- "C:/Users/jonghoon.kim/Dropbox/Typhoid mapping/country_data/" 
cntries <-
  c("Angola", "Benin", "Botswana", "Burkina Faso", "Burundi", "Cameroon",
   "Central African Republic", "Chad", "Congo", "Cote d'Ivoire",
   "Eritrea", "Ethiopia", "Gabon", "Gambia", "Ghana", "Guinea", "Guinea-Bissau",
   "Kenya", "Lesotho", "Liberia", "Madagascar", "Malawi", "Mali", "Mauritania",
   "Mauritius", "Mozambique", "Namibia", "Niger", "Nigeria", "Rwanda", "Senegal",
   "Sierra Leone", "South Africa", "Tanzania", "Togo", "Uganda",
   "Zaire", "Zambia", "Zimbabwe")

datlist <- list()             
for(i in 1:length(cntries)) {
# for(i in 9:9) {
  cntry <- cntries[i]
  d <- read_excel(paste0(path_country_data, cntry, "/", cntry,".xlsx"))
  cat("i =", i, ", country =", cntry, 
      ", nrow =", nrow(d), ", ncol =", ncol(d), "\n")
  datlist[[i]] <- d
}
# 
# refnm <- names(datlist[[1]])
# for(i in 1:length(datlist)) {
#   nm <- names(datlist[[i]])
#   cat("i=", i, "\n")
#   if(!identical(refnm, nm)) {
#     message(which(refnm != nm))
#   }   
# }

dat <- do.call("rbind", datlist)
nms <- names(datlist[[1]])

# remove NA rows
dat <- dat[!(sum(is.na(dat[,1:ncol(dat)])) == ncol(dat)), ]
# remove INCLUDE == 2, which indicates studies that were not included in the new search
# > dim(dat)
# [1] 3648   21
dat <- dat[dat$INCLUDE == 0 | dat$INCLUDE == 1, ]
# > dim(dat)
# [1] 2377   21
# remove studies with diagnostic methods are missing
dat <- dat[!is.na(dat$DIAGNOSTIC_METHODS) & dat$DIAGNOSTIC_METHODS != "", ]
# > dim(dat)
# [1] 706  21
# dstamp <- format(Sys.time(), "%Y%m%d")
# data.table::fwrite(dat, paste0("out/occ_data_", dstamp, ".csv"))
# dat <- data.table::fread(paste0("out/occ_data_", dstamp, ".csv"))
```

### Clean diagnostic methods
```{r}
blood_culture_synonyms <- 
  c("urine, blood, or stool culture",
    "blood or stool culture, or Widal-Felix",
    "blood/urine/stool culture",
    "urine/blood /stool culture",
    "stool/blood culture",
    "blood culture, stool culture, or serology",
    "blood culture and PCR",
    "blood culture or fever with stool culture",
    "antibody tetsing or blood/stool culture",
    "blood and stool culture","blood culture",
    "Blood culture", "blood culture and serology",
    "blood culture or PCR",
    "blood culture, Widal, or fecal cultuure",
    "blood isolates",
    "blood or stool culture",
    "blood or stool culture perforation",
    "blood, stool, and urine culture",
    "blood, stool, or urine culture",
    "blood, stool, urine culture, or Widal test",
    "blood/stool",
    "blood/stool culture also clinical symptoms",
    "blood/stool/widal",
    "blood/urine/stool",
    "Confirmed cases with positive blood culture",
    "Blood culture, Widal test, stool culture",
    "blood/stool culture",
    "blood or stool culture, or Widal-Felix",
    "isolate from blood, stool, or urine",
    "blood culture and widal test",
    "blood culture and agglutination test",
    "blood culture and Widal",
    "blood culture/widal test",
    "blood or stool culture, or Widal",
    "Widal test and blood culture",
    "clinical, blood culture",
    "blood and bile culture",
    "blood culture and intestinal perforation",
    "isolation from blood, stool, or urine culture",
    "stool and blood culture",
    "blood/stool culture",
    "Blood culture, Widal test, stool culture",
    "blood culture followed by PCR",
    "blood culture, Widal test, stool culture",
    "blood, urine culture")

bone_marrow_culture_synonyms <- 
  c("blood, stool or bone marrow culture",
    "blood and bone marrow cultures",
    "blood or bone marrow culture",
    "blood and bone marrow culture",
    "blood, bone marrow aspirate, stool, and urine cultures",
    "blood, bone marrow, stool, or urine culture")

CSF_culture_synonyms <- 
  c("Cerebrospinal fluid culture",
    "spinal fluid culture",
    "isolation from CSF, blood, or stool",
    "CSF culture",
    "blood or CSF culture",
    "fecal, blood, urine, CSF",
    "blood, stool, urine, aspirates, cerebrospinal fluid, wounds, and abscesses",
    "blood, CSF, urine",
    "Culture of blood, CSF, and joint",
    "Culture of blood, CSF, or pleural fluid")

aspirate_culture_synonyms <- 
  c("Aspirates culture",
  "culture of pus",
  "Wound culture",
  "wound swab culture",
  "Ear and throat culture",
  "Sputum culture",
  "rectal swab specimens examined bacteriologically") 


Widal_test_synonyms <- 
  c("Widal test/typhoid ileal perforation",
    "Widal, perforation",
    "perforation, Widal",
    "agglutination and titration (> 4-fold)", 
    "agglutination test",
    "clinical and Widal test",
    "clinical/ widal",
    "typhoid perforation/widal test",
    "widal",
    "Widal",
    "Widal-Felix",
    "widal test",
    "Widal test",
    "Widal test (Widal titres >= 160 to 0 antigen)",
    "perforation, Widal, PCR of Salmonella Typhi DNA",
    "blood smear, widal test",
    "clinical or Widal",
    "perforation and Widal test",
    "Widal and Weilfelix direct card agglutination tests (DCAT)",
    "Widal and Weilflex of direct card agglutination test (DCAT) and tube based titration tests",
    "ileal perforation, Widal test",
    "Tube agglutination test",
    "clinical/widal")

autopsy_synonyms <- c("post morten report",
                      "autopsy",
                      "Autopsy",
                      "coroner's autopsies",
                      "autopsy",
                      "post mortem")

culture_synonyms <- 
  c("lab confirmed",
    "laboratory confirmed",
    "confirmed",
    "Isolation of Salmonella typhi",
    "culture",
    "isolated",
    "bacteriologically",
    "bacterial isolation",
    "Culture",
    "MDR isolates",
    "culture confirmed",
    "isolate (drug resistance)",
    "microbiological isolates",
    "bacterial isolation",
    "biobanked isolates cultured",
    "clinical or culture",
    "culture",
    "culture positive",
    "culture, Widal test",
    "cultured",
    "doctor diagnosed and self diagnosed",
    "isloation",
    "isolate",
    "isolated pathogen(blood culture)",
    "isolates",
    "isolates (antigen detected)",
    "isolation",
    "isolation (just written human sample)",
    "S.Typhi isolates",
    "salmonella strain",
    "isolated",
    "suspected and confirmed cases",
    "microbiological diagnostic tests (rapid tests, serologies, cultures and molecular analyses)", 
    "Salmonella typhi strain",
    "bacteremic invasive S.typhi",
    "Widal test or bacteriologically confirmed",
    "bacteriological",
    "Culture proven(not specific)/widal test",
    "culture and Widal test",
    "bacteria isolated(exact method not given)",
    "bacteriologically",
    "Isolation and PCR",
    "S. typhi isolation",
    "isolation of Salmonella typhi",
    "microbiologically confirmed")

NA_synonyms <- 
  c("Survey questionnaire",
    "blood",
    "blood test",
    "carrier",
    "extensive diagnostic",
    
    "NA",
    "not clearly mentioned",
    "not given",
    "online survey",
    "recorded data",
    "missing",
    "history of typhoid",
    "besides standard measure lumbar punture & CSF",
    "surveillance data used",
    "National Health Laboratory Services data",
    "previous data",
    "maybe clinical not clear in the article",
    "standard methods with blood samples",
    "history based on self report",
    "questionaire",
    "report at medical record",
    "self reported")

biochemical_synonyms <- 
  c("biochemical",
    "biochemical testing of urine sample",
    "biochemical tests of environmental samples",
    "biochemical tests of environmental samples")

rapid_test_synonyms <- c("fever and positive rapid typhoid test",
                         "Blood-RDT (Typhoid IgM)",
                         "blood-RDT (Typhoid IgM)",
                         "rapid diagnostic test",
                         "rapid diagnostic test (IgM)",
                         "rapid_test",
                         "RDT")

clinical_synonyms <- 
  c("travel or tropical medicine clinics record",
    "Retrospective review of death records",
    "cases reported",
    "typhoid fever",
    "cause of death",
    "Typhoid perforation",
    "Intestinal perforation",
    "clinically", 
    "clinically diagnosed",
    "clinical",
    "clinical diagnosis",
    "digestive perforation",
    "typhoid gall bladder perforation",
    "gallbladder perforation",
    "cholecystitis", 
    "clinical and blood test",
    "clinical and radiological findings",
    "clinical symptoms",
    "hospital record",
    "hospital record( through clinical classifications software)",
    "ileal perforation",
    "ileal performation, Widal test",
    "intestinal perforation",
    "medical record(clinical)",
    "microbiologically proven",
    "multiple jejunal perforations and positive serology",
    "perforated enteritis",
    "perforation",
    "peritonitis",
    "peritonitis/typhoid fever",
    "self reported (clinical symptoms)",
    "sepsis",
    "typhoid ileal perforation",
    "typhoid intestinal perforation",
    "typhoid perforation",
    "clinical and serological criteria",
    "suspected typhoid cases",
    "suspected cases",
    "Intestinal Perforation",
    "A suspected typhoid case was defined as diagnosis of IP or symptoms of fever, abdominal pain, and one or more of the following: vomiting, diarrhea, constipation, joint pain, headache, general body weakness, clinical suspicion of IP, or failure to respond to antimalarials",
    "small bowel perforation",
    "histology",
    "peforation or other clinical signs",
    "enterocutaneous fistula",
    "clinical features and laboratory investigations",
    "demographic, clinical, and operative details",
    "septicaemia",
    "clinical or laboratory evidence",
    "clinical presentation",
    "perforated typhoid ileitis",
    "typhoid enteritis",
    "intestinal typhoid perforation(bowel resection)",
    "clinical (hospital record)",
    "case files",
    "medical record/folios",
    "intestinal typhoid perforation (bowel resection)")

environmental_sample_culture_synonyms <- 
  c("environmental sample culture",
   "S. typhi isolated from water")

stool_culture_synonyms <- 
  c("stool",
    "stool culture",
    "Widal and stool culture",
    "stool_culture",
    "stool culture (carrier state)",
    "Widal test & stool culture",
    "Stool culture",
    "stool_culture",
    "stool culture, extensive diagnostic")

urine_culture_synonyms <- 
  c("urine culture",
    "Urine culture",
    "urine, Widal test",
    "urine_culture",
    "Urine")

serology_synonyms <- 
  c("serology",
    "antibody titre",
    "antibody titre",
    "positive titre",
    "clinical diagnosis and serology",
    "serologically diagnosed",
    "clinical and serological criteria") 

PCR_synonyms <- 
  c("PCR",
    "genome sequence",
    "TaqMan Array Card",
    "perforation, Widal, PCR of Salmonella Typhi DNA")

dat$diagnostic_method <- dat$DIAGNOSTIC_METHODS
# NA assignment must be before blood assignment.
# Otherwise, blood will be replaced with NA
dat$diagnostic_method[dat$diagnostic_method %in% NA_synonyms] <- "Unclear"
dat$diagnostic_method[dat$diagnostic_method %in% PCR_synonyms] <- "PCR" 
dat$diagnostic_method[dat$diagnostic_method %in% bone_marrow_culture_synonyms] <- "Bone marrow culture" 
dat$diagnostic_method[dat$diagnostic_method %in% CSF_culture_synonyms] <- "CSF culture" 
dat$diagnostic_method[dat$diagnostic_method %in% aspirate_culture_synonyms] <- "Aspirate culture" 
dat$diagnostic_method[dat$diagnostic_method %in% serology_synonyms] <- "Serology" 
dat$diagnostic_method[dat$diagnostic_method %in% blood_culture_synonyms] <- "Blood culture" 
dat$diagnostic_method[dat$diagnostic_method %in% Widal_test_synonyms] <- "Widal test"
dat$diagnostic_method[dat$diagnostic_method %in% autopsy_synonyms] <- "Autopsy"
dat$diagnostic_method[dat$diagnostic_method %in% culture_synonyms] <- "Culture"
dat$diagnostic_method[dat$diagnostic_method %in% biochemical_synonyms] <- "Biochemical test"
dat$diagnostic_method[dat$diagnostic_method %in% rapid_test_synonyms] <- "Rapid test"
dat$diagnostic_method[dat$diagnostic_method %in% clinical_synonyms] <- "Clinical symptoms"
dat$diagnostic_method[dat$diagnostic_method %in% environmental_sample_culture_synonyms] <- "Environmental sample culture"
dat$diagnostic_method[dat$diagnostic_method %in% stool_culture_synonyms] <- "Stool culture"
dat$diagnostic_method[dat$diagnostic_method %in% urine_culture_synonyms] <- "Urine culture"
unique(dat$diagnostic_method)                      
# table(dat$diagnostic_method)

df <- data.frame(matrix(NA, nrow=16, ncol=2))
names(df) <- c("Category", "Actual terms used in the literature")

df$Category <- c("Aspirate culture", "Autopsy", "Biochemical test", "Bone marrow culture", "Blood culture", "Clinical symptoms", "CSF culture", "Culture", "Environmental sample culture", "Unclear", "PCR", "Rapid test", "Serology", "Stool culture", "Urine culture", "Widal test")

strs <- list(aspirate_culture_synonyms, autopsy_synonyms, biochemical_synonyms, bone_marrow_culture_synonyms, blood_culture_synonyms, clinical_synonyms, CSF_culture_synonyms, culture_synonyms, environmental_sample_culture_synonyms, NA_synonyms, PCR_synonyms, rapid_test_synonyms, serology_synonyms, stool_culture_synonyms, urine_culture_synonyms, Widal_test_synonyms)

# strs2 <- sapply(1:length(strs), function(z) toString(sprintf('"%s"', strs[[z]])))
strs2 <- sapply(1:length(strs), function(z) toString(sprintf('"%s"', strs[[z]])))

df$`Actual terms used in the literature` <- strs2

# fwrite(df, "out/diagnostic_method_category_phrases.csv")
library(dplyr)
d1 <- fread("out/study_by_diagnostic_method.csv")

# d2 <- fread("out/diagnostic_method_category_phrases.csv", quote="")
# read and write again causes double double quotes
d <- merge.data.table(d1, df, by.x = "Diagnostic method", by.y = "Category")
fwrite(d, "out/diagnostic_method_freq_and_terms.csv")
```

### Clean study types
```{r}
case_report_synonyms <- 
  c("Reports of sporadic cases",
    "Report of sporadic cases",
    "Case report",
    "sporadic case",
    "case report",
    "report of sporadic cases",
    "case series",
    "reports of sporadic cases",
    "report",
    "reports of spaoradic cases")

retrospective_synonyms <- 
  c("retrospective study",
    "retrospective",
    "review of hospital record",
    "review of hospital records",
    "review of sporadic cases",
    "Review of  hospital record",
    "Review of  hospital record (cross sectional study)",
    "Review of hospital record",
    "hospital record",
    "Retrospective data analysis",
    "review of hospital data",
    "Review of hospital data",
    "reviewof hopsital record",
    "review of hospital rec",
    "review of hospital  record",
    "review   of hospital record",
    "reviw of hospital record",
    "retrospective review of hospital record",
    "review of hospitalrecord",
    "lab record",
    "review of hospital records (retrospective study of biobanked samples)")

prospective_synonyms <- 
  c("longtudinal surveillance",
    "longitudinal surveillance (hospital based)",
    "prospective study",
    "Longitudinalstudy",                    
    "longitudinal study",
    "Longitudinal surveillance - IR",
    "longitudinal surveillance - IR",
    "longitudinal surveillance",
    "Prospective study",
    "Prospective record",
    "Prospective study (Cross sectional study)",
    "Prospective study and review of hospital record",
    "prospective and case control study",
    "prospective clinical study",
    "prospective and case-controlled study",
    "prospective  study",
    "review of hospital record  and prospective study",
    "logitudinal surveilance",
    "hospital record (retrospective) and longitudinal study",
    "review of hospital record followed by prospective study",
    "longitudinal study(hospital record)",
    "longitudinal surveillance(hospital record)",
     "Longitudinal surveillance")

cross_sectional_synonyms <- 
  c("Cross-sectional survey(hospital based)",
    "Cross sectional study",
    "cross sectional study",
    "cross-sectional multicentric study",
    "cross-sectional study",
    "Cross-sectional survey",
    "review of hospital record (Cross-sectional study)",
    "cross-sectional")

case_control_synonyms <- 
  c("case control study",
    "Case control study",
    "Case Control study",
    "Case -Control study",
    "case-control study")

literature_review_synonyms <- 
  c("Literature review",
    "literature review")

outbreak_synonyms <- c("outbreak",
                       "Outbreak investigation",
                       "Outbreak Investigation",
                       "oubreak",
                       "outbreak investigation",
                       "hospital record/outbreak" ) 

clinical_trial_synonyms <- c("clinical trial")

dat$study_type <- dat$STUDY_TYPE
dat$study_type[dat$study_type %in% case_report_synonyms] <- "Case report"
dat$study_type[dat$study_type %in% retrospective_synonyms] <- "Retrospective study"
dat$study_type[dat$study_type %in% prospective_synonyms] <- "Prospective study"
dat$study_type[dat$study_type %in% cross_sectional_synonyms] <- "Cross-sectional study"
dat$study_type[dat$study_type %in% case_control_synonyms] <- "Case-control study"
dat$study_type[dat$study_type %in% literature_review_synonyms] <- "Literature review"
dat$study_type[dat$study_type %in% outbreak_synonyms] <- "Outbreak"
dat$study_type[dat$study_type %in% clinical_trial_synonyms] <- "Clinical trial"

unique(dat$study_type)
datin <- dat[dat$INCLUDE == 1,]
unique(dat[dat$INCLUDE == 1,]$study_type)

# dstamp <- format(Sys.time(), "%Y%m%d")
# data.table::fwrite(dat, paste0("out/occ_data_", dstamp, ".csv"))
# dat <- data.table::fread(paste0("out/occ_data_", dstamp, ".csv"))

```

### Number of studies by diagnostic method
```{r}
library(dplyr)
library(data.table)
dat <- fread("data/occ_data_20221023_edit.csv")
# There are duplicates because of studies that hold multiple sub-national locations
# They need to be adjusted to calculate the number of studies

# dat %>% distinct(STUDY_INFO, .keep_all = TRUE) -> dat_2
dat_ <- dat[!duplicated(dat$STUDY_INFO),]

dat_ %>% 
  filter(DIAGNOSTIC_METHOD %in% c("aspirate", "blood", "bone marrow", "CSF", "urine")) %>% 
  summarize(n = n())
dat_ %>%
  filter(DIAGNOSTIC_METHOD %in% c("autopsy", "biochemical", "clinical", "culture", "environmental", "PCR", "rapid", "serology", "stool", "Widal")) %>% 
  summarize(n = n())

tb <- table(dat_$DIAGNOSTIC_METHOD)  

totstudy <- nrow(dat_)
dat_ %>% 
  group_by(DIAGNOSTIC_METHOD) %>%
  summarise(cnt = n(), percent = 100 * cnt / totstudy) -> d
d <- d[order(d$cnt, decreasing = T),]

df <- data.frame(matrix(NA, nrow = nrow(d), ncol=2))
names(df) <- c("Diagnostic method", "Frequency [n (%)]")
df$`Diagnostic method` <- d$DIAGNOSTIC_METHOD
df$`Frequency (n, %)` <- paste0(d$cnt, " (", format(round(d$percent, digits=2), big.mark=",", trim=TRUE), ")") 
# fwrite(df, "out/study_by_diagnostic_method.csv")
```

### Diagnostic method - allow multiple methods for one study
```{r}
dat_ <- fread("data/occ_data_20221023_edit.csv")

x <- dat_$DIAGNOSTIC_METHOD
bc <-grepl("blood", x, ignore.case = T) & !grepl("RDT", x, ignore.case = T)
stool <- grepl("stool", x, ignore.case = T)
urine <- grepl("urine", x, ignore.case = T)
sero <- grepl("serolog", x, ignore.case = T) | grepl("antibody", x, ignore.case = T)
csf <- grepl("CSF", x, ignore.case = T) | grepl("cerebro", x, ignore.case = T)
bm <- grepl("bone marrow", x) 
rdt <- grepl("RDT", x, ignore.case = T) | grepl("rapid", x, ignore.case = T)
widal <- grepl("widal", x, ignore.case = T) | grepl("agglutin", x, ignore.case = T)
clin <- grepl("clinical|perfor|cholecystitis|hospital|medical|peritonitis|septicaemia|histology|case files", x, ignore.case = T)
autop <- grepl("post morten|autops", x, ignore.case = T)

nrow(dat_[bc,])
nrow(dat_[csf,])
nrow(dat_[bm,])
nrow(dat_[stool,])
nrow(dat_[urine,])
nrow(dat_[sero,])
nrow(dat_[widal,])
nrow(dat_[clin,])
nrow(dat_[autop,])
nrow(dat_[rdt,])

```

### Number of studies by study type
```{r}
table(dat_$study_type)
# let's include studies that is INCLUDE == 1
dat__ <- dat_[dat_$INCLUDE == 1, ]
dat__ %>% 
  group_by(study_type) %>% 
  count() -> st
```

### Studies that provided incidence rates

```{r}
IR_study_synonyms <- c("Longitudinal surveillance - IR",
                       "longitudinal surveillance - IR")
IR_study <- dat__[dat__$STUDY_TYPE %in% IR_study_synonyms, ]
IR_study

prospective_study <- 
  dat__[dat__$STUDY_TYPE %in% IR_study & 
          dat__$diagnostic_method %in% 
          c("aspirate", "blood", "bone marrow", "CSF", "urine"), ]
```

### Sub-national regions
Sub-national regions were cleaned by reviewing the Excel sheets

```{r}
adm0 <- unique(dat__$ADM_0)
library(dplyr)
dat__ %>% 
  group_by(ADM_0) %>%
  count(SUB_NATIONAL_REGION) -> dat_subnational

dat_ %>% 
  filter(ADM_0 == "Nigeria") %>%
  count(SUB_NATIONAL_REGION) -> nga_sub

dat_[grepl("Univ. of Ilorin", dat_$SUB_NATIONAL_REGION), ]

# 1. Wesley Guild Hospital Ilesha (Ilesha city, Osun State, southwest Nigeria), 2. Obafemi Awolowo University Teaching Hospital(Ife city, Osun State)
# 3. University Teaching Hospital, Ado-Ekiti, Ekiti State -> Ekiti State University Teaching Hospital
# 4. University Teaching Hospital, Calabar -> University of Calabar Teaching Hospital, Calabar city, Southern Nigeria (Calabar is a port city in southern Nigeria)
# 5. University of Nigeria, Nsukka District, of Enugu State - > University of Nigeria, Nsukka District, Enugu State 
# 6. University of Port Harcourt Teaching Hospital Port Harcourt -> University of Port Harcourt Teaching Hospital, Port Harcourt (Capital and largest city of the state), Rivers State
# 7. University of Nigeria Teaching Hospital, Ituku,Ozalla, Enugu State, South-East Nigeria
# University of Nigeria Teaching Hospital, Enugu (capital), Enugu State, Southeastern Nigeria
# University of Nigeria Teaching Hospital, Ituku, Ozalla
# 8. University of Lagos, Idi-Araba, Lagos
# 9. University of Maiduguri Teaching Hospital, Borno State
# University of Maiduguri Teaching Hospital, Maiduguri (the capital and the largest city), Borno State, Northeastern Nigeria
# 10. University of Ibadan, Ibadan
# 11. University of Ilorin Teaching Hospital, Ilorin
# 12. University of Benin Teaching Hospital, Benin City
# University of Calabar Teaching Hospital, Calabar
# 13. University College Hospital, Ibadan, Nigeria
# 14. University of Ado-Ekiti Health Centre
# 15. University College Hospital, Catholic Hospital Oluyoro (Oke Ofa-Oje, Ibadan, Nigeria), 16. Adeoyo Maternity Teaching Hospital(Yemetu St, 200285, Ibadan, Nigeria), 17. General Hospital Moniya (Old Oyo Rd, 200136, Ibadan, Nigeria), 18. Ring Road State Hospital and Health Centres at Sango.
# University College Hospital, Ibadan
# 19. Olabisi Onabanjo University Teaching Hospital, Sagamu, Ogun State
# 20. Otukpo General Hospital (6544+23Q, 972261, Otukpo, Nigeria)
# Paediatric Surgery Division, Department of Surgery, Olabisi Onabanjo University Teaching Hospital, Sagamu
# Obafemi Awolowo University Teaching Hospitals Complex Ile-Ife, Nigeria,
# Obafemi Awolowo University, Ile-ife, Nigeria
# Obafemi Awolowo University Teaching Hospital Complex, Ile-Ife
# 21. Nigerian Institute of Medical Research, Yaba, Lagos
# 22. Nnamdi Azikiwe Teaching Hospital, Nnewi, Nigeria
# 23. Nnamdi Azikiwe University, Awka, Anambra State
# 24. Lagos University Teaching Hospital
# 25. Landmark University Medical Center, Omuaran
# 26. Leadeks Medical Centre, Benim City Edo State, Nigeria
# 27. Maiduguri Teaching Hospital
# 28. Ladoke Akintola University of Technology Teaching Hospital, Osogbo, Osun State
# 29. Jos University Teaching Hospital, Jos
# 30. Imo State University, Owerri
# 31. General Hospital Abudu, Edo State, Nigeria
# 32. General hospital, Ikare Akoko, Ondo state
# 33. Federal Medical Center (FMC) Birnin Kudu
# 34. Federal Medical Centre (FMC), Azare
# 35. Federal Medical Centre Makurdi
# 36. College of Education Health Clinic, Akwanga
# 37. Ekiti State Specialist Hospital, Children's Ward, Ado-Ekiti, Ekiti State
# Ekpoma, Edo State
# 38. Enugu State University of Science and Technology Teaching Hospital
# 39. Benin teaching Hospital (UBTH), Benin City
# 40. Aminu Kano Teaching Hospital, 41. Hasiya Bayero Pediatric Hospital (HBPH), 42. Murtala Muhammad Specialist Hospital (MMSH), Kano
# 43. Aminu Kano Teaching Hospital, Kano
# Aminu Kano University Teaching Hospital, Kano
# 44. Baptist Medical Centre in Ogbomoso, Nigeria (BMCO)
# Baptist Medical Centre, Ogbomoso
# 45. Bayero University, Kano; 46. Ahmadu Bello University, Zaria; 47.Kaduna State University, Kaduna; University of Maiduguri, Borno; 48. Usman Danfodio University, Sokoto; 49. Gombe State University, Gombe; University of Ilorin, Kwara; and University of Jos, Plateau
# 50. Bells University of Technology and Covenant University Health Centers, Ota, Ogun State
# Aminu Kano Teaching Hospital (AKTH), Hasiya Bayero Pediatric Hospital and Murtala Specialist Hospital, Kano
# 51. Ahmadu Bello University Teaching Hospital, Zaria
# Ahmadu Bello University Teaching Hospital, Zaria, Nigeria.
# 52. Akaraugo Hospital in Owerri, Imo state
# 53. Artillery Brigade Medical Centre, Minna
# A community health centre in Abeokuta
# 54. Abeokuta and Minna
# 55. Abia State University Teaching Hospital, Aba, Abia State
```


### Points in which over 20 km by 20 km were identified

```{r}
# datin_lg <- datin[grepl("km", datin$NOTES),]
# # extract numbers before km
# nt <- datin_lg$NOTES[1]
# # gsub(".*([0-9]+).*", "\\9", nt)
# nt <- "Approx. 15 km by 20 km"
# 
# 
# library(stringr)
# 
# for (i in 1:nrow(datin_lg)) {
#   wd <- datin_lg$NOTES[i]
#   # cat(wd, "\n")
#   # x <- str_extract_all(wd, regex("\\(?[0-9]+\\)?"))[[1]]
#   x <- str_extract_all(wd, regex("\\d+(?= km)|\\d+(?=km)"))[[1]]
#   
#   # x <- str_extract_all(wd, regex("\\d+(?= km)|\\d+(?=km)"))[[1]]
#   if (any(as.numeric(x) > 20)){
#     cat("extracted:", i, ", ", x, "\n")  
#   }
#   # invisible(readline(prompt="Press [enter] to continue"))
# }

datin$gridx <- rep(NA, nrow(datin))
datin$gridy <- rep(NA, nrow(datin))
datin$catchmentx <- rep(NA, nrow(datin))
datin$catchmenty <- rep(NA, nrow(datin))

for (i in 1:nrow(datin)) {
  wd <- datin$NOTES[i]
  # cat(wd, "\n")
  if (!is.na(wd)) {
    x <- as.numeric(str_extract_all(wd, regex("\\d+(?= km)|\\d+(?=km)"))[[1]])
    # if (any(as.numeric(x) > 20)) {
    #   cat("extracted:", i, ", ", x, "\n")  
    # }
    # if (length(x) > 2) {
    #   cat("extracted:", i, ", ", x, "\n")  
    # }
    # if (length(x) < 2) {
    #   cat("extracted:", i, ", ", x, "\n")  
    # }
    if (length(x) >= 2 & any(as.numeric(x) > 20)){
      cat("extracted:", i, ", ", x[1], x[2],"\n")
      datin$gridx[i] <- round(x[1]/20)
      datin$gridy[i] <- round(x[2]/20)
      datin$catchmentx[i] <- x[1]
      datin$catchmenty[i] <- x[2]
    }
  }
}

# dstamp <- format(Sys.time(), "%Y%m%d")
# data.table::fwrite(datin, paste0("out/occ_data_inc_", dstamp, ".csv"))
```

### Study frequency by country and year
```{r}
source("util/util.R")
library(data.table)
library(readxl)
# d <- read_xlsx("data/occ_data_20221023.xlsx")
d <- fread("data/occ_data_20221025.csv")

dt <- setDT(d)
dt <- dt[INCLUDE == 1]
dt <- dt[, c("STUDY_INFO", "STUDY_DATE_BEGIN", "STUDY_DATE_END", "NUM_CASE", "SUB_NATIONAL_REGION", "AREA_SIZE", "YEAR_PUBLICATION", "LONGITUDE", "LATITUDE", "AREA_HOSPITAL", "COUNTRY", "DIAGNOSTIC_METHOD2", "STUDY_TYPE2")]
names(dt) <- c(names(dt)[1:11], "DIAGNOSTIC_METHOD", "STUDY_TYPE")
dt$COUNTRY <- clean_country_names(dt$COUNTRY)
dt <- dt[YEAR_PUBLICATION >= 2000]
# fwrite(dt, "data/occ_data_20221023_edit.csv")

library(dplyr)
unique(dt$COUNTRY)

dt %>% 
  distinct(STUDY_INFO, .keep_all = TRUE) %>% 
  filter(YEAR_PUBLICATION >= 2000) -> dt2

dt2 %>%
  group_by(YEAR_PUBLICATION) %>%
  summarise(n=n()) -> cnt
# plot(cnt$YEAR_PUBLICATION, cnt$n)
plot(cnt$YEAR_PUBLICATION, cnt$n)

library(ggplot2)
p <- ggplot(cnt, aes(YEAR_PUBLICATION, n)) + 
  geom_point() + 
  geom_line() + 
  theme_bw() +
  scale_x_continuous(sec.axis = dup_axis()) +
  scale_y_continuous(sec.axis = dup_axis()) + 
  theme(axis.ticks.length=unit(-2.75, "pt"), 
        axis.text.x.top = element_blank(),      
        axis.text.y.right = element_blank(),  
        axis.title.x.top = element_blank(), 
        axis.title.y.right = element_blank()) +
  labs(x = "Year of publication", y = "Number of articles")

ggsave(paste0("plots/study_freq_yr_", tstamp(), ".png"), p, 
       width = 3.4, height = 2.7, units="in") 

dt2 %>% 
  group_by(COUNTRY, YEAR_PUBLICATION) %>%
  summarize(count=n()) -> d_co_yr

dt2 %>%
  group_by(YEAR_PUBLICATION) %>%
  summarize(count = n()) -> d_tot

d_tot$COUNTRY <- "Total"
d_co_yr <- bind_rows(d_co_yr, d_tot)
cntry_ordered <- unique(d_co_yr$COUNTRY)
d_co_yr$COUNTRY <- factor(d_co_yr$COUNTRY, levels = rev(cntry_ordered))
# d_co_yr$COUNTRY <- factor(d_co_yr$COUNTRY)
cntry_labels <- cntry_ordered
cntry_labels[6] <- "CAR" # abbreviation for better displaying
cntry_labels[25] <- "Tanzania" # abbreviation for better displaying
# tiff("figs/reports_co_yr.tiff", units="in", width=5.4, height=7, res=300)

p <- ggplot(data = d_co_yr) + 
  geom_point(aes(x = YEAR_PUBLICATION, y = COUNTRY, size = count)) +
  # geom_point(aes(x = year_pub, y = country, size = count_lab)) +
  scale_size_continuous(limits = c(1, 50)) +
  scale_x_continuous(breaks = seq(2000, 2020, by=5), limits = c(2000, 2020)) +
  # scale_x_continuous(sec.axis = dup_axis()) + 
  scale_y_discrete(breaks = cntry_ordered, labels = cntry_labels) + 
  # scale_y_discrete(sec.axis = dup_axis()) + 
  labs(x="Year of publication", y = NULL, size="No. of reports" ) +
  theme_bw() + 
  theme(axis.ticks.length=unit(-2.75, "pt"), 
        legend.position = "bottom")
# p
fac <- 1.8
ggsave(paste0("plots/study_freq_yr_cntry_", tstamp(), ".png"), p, 
       width = 3.4*fac, height = 2.7*fac, units="in") 


dt2 %>% group_by(COUNTRY) %>%
  summarise(min_year = min(YEAR_PUBLICATION)) -> dminyear 

head(dminyear[order(dminyear$min_year, decreasing = T),], n=20)
```



### Plot the occurrence points on the map
Find the grid cells where occurrence points fall 

```{r}
library(raster)
library(readxl)
library(data.table)
source("util/map_functions.R")
source("util/ggplot2_theme.R")
source("util/util.R")

rst <- readRDS("data/covariates/prediction/elevation_20km_africa.rds")
afss <- readRDS("data/africa_sub_Sahara_adm0_shp.rds")
af_adm0 <- readRDS("data/africa_adm0_shp.rds")
# occ <- read_xlsx("data/occ_data_20221023.xlsx")

d <- read_xlsx("data/occ_data_20221023.xlsx")
dt <- setDT(d)
dt <- dt[INCLUDE == 1]
dt <- dt[, c("STUDY_INFO", "STUDY_DATE_BEGIN", "STUDY_DATE_END", "NUM_CASE", "SUB_NATIONAL_REGION", "AREA_SIZE", "YEAR_PUBLICATION", "LONGITUDE", "LATITUDE", "AREA_HOSPITAL", "COUNTRY", "DIAGNOSTIC_METHOD2", "STUDY_TYPE2")]

dt$COUNTRY <- clean_country_names(dt$COUNTRY)
 
# names(dt) <- c(names(dt)[1:11], "DIAGNOSTIC_METHOD_SUMMARY", "STUDY_TYPE")
dt$STUDY_DATE_BEGIN <- round(as.numeric(dt$STUDY_DATE_BEGIN), digits = 0)
dt$STUDY_DATE_END <- round(as.numeric(dt$STUDY_DATE_END), digits = 0)
# # dt %>% distinct(STUDY_INFO) %>% summarise(cnt = n())
# dt2 <- dt[YEAR_PUBLICATION >= 2000]
dt3 <- as.data.frame(dt2)
for (i in 1:ncol(dt3)) {
  dt3[,i] <- ifelse((is.na(dt3[,i]) | dt3[,i] == "NA"), "NA", dt3[,i])
}
ext = str_extract_all(dt3$AREA_SIZE, "[0-9]+.km", simplify = TRUE)
# > dim(ext)
# [1] 386   5
# > View(ext)
# > which(ext[,3] != "")
# [1] 372
# row 372 checked manually and we can use the first two columns
dt3$AREA_SIZE <- NA
for (i in 1:nrow(dt3)) {
  dt3$AREA_SIZE[i] <- ifelse(sum(ext[i,1:2] == "") == 0, paste(ext[i,1:2], collapse = " by "), "NA")
}
#"100km by 50 km" to "100 km by 50 km" # manually
dt3$AREA_SIZE[313] <- "100 km by 50 km"

fwrite(dt3, paste0("data/occ_data_", tstamp(), ".csv"))
# manual edit to remove any additional word in the AREA_SIZE column

# fwrite(dt2, paste0("data/occ_data_", tstamp(), "_edit.csv"))
# # dt2 %>% distinct(STUDY_INFO) %>% summarise(cnt = n())
nms <- names(dt2)
# dryad (https://datadryad.org/stash/resources/202931/upload) submission system appears to have file checking, which may have recognized that columns 2 and 3 have "DATE" in the name but the contents are integers.  Hence column names were changed. 
names(dt2) <- c(nms[1], "STUDY_YEAR_BEGIN", "STUDY_YEAR_END" ,names(dt2)[4:length(nms)])
# renaming did not solve the issue and therefore, changed the columns to characters like other columns
dt2$STUDY_DATE_BEGIN <- as.character(dt2$STUDY_DATE_BEGIN)
dt2$STUDY_DATE_END <- as.character(dt2$STUDY_DATE_END)

fwrite(dt2, paste0("data/occ_data_", tstamp(), "_edit.csv"))

dculture <- dt2[grepl("culture", DIAGNOSTIC_METHOD2, ignore.case = TRUE)]
dother <-  dt2[!grepl("culture", DIAGNOSTIC_METHOD2, ignore.case = TRUE)]


rst <- mask(rst, afss)
rst_ <- rst
rst <- raster::setValues(rst, 0)
## culture-based occurrence
library(dplyr)
dculture %>% 
  filter(!is.na(LONGITUDE)) %>% 
  distinct(LONGITUDE, .keep_all = TRUE) -> dculture2

occmat <- data.frame(X = as.double(dculture2$LONGITUDE),
                     Y = as.double(dculture2$LATITUDE))
occmat <- occmat[!is.na(occmat$X) & !is.na(occmat$Y), ]
for (i in 1:nrow(occmat)){
  # cells <- cellFromPolygon(rmask, shp)
  cells <- raster::extract(rst, occmat[i,], df = TRUE, cellnumbers = TRUE)
  values(rst)[unlist(cells$cell)] <- 1 
}

dother %>%
  mutate(LONGITUDE = as.double(LONGITUDE), LATITUDE = as.double(LATITUDE)) %>% 
  filter(!is.na(LONGITUDE) & !is.na(LATITUDE)) %>% 
  distinct(LONGITUDE, .keep_all = TRUE) -> dother2

occmat2 <- data.frame(X = dother2$LONGITUDE, Y = dother2$LATITUDE)
occmat2 <- occmat2[!is.na(occmat2$X) & !is.na(occmat2$Y), ]
for (i in 1:nrow(occmat2)){
  cells <- raster::extract(rst, occmat2[i,], df = TRUE, cellnumbers = TRUE)
  values(rst)[unlist(cells$cell)] <- 2 
}

rst <- mask(rst, afss)
rpts <- rasterToPoints(rst)
rptsdf <- as.data.frame(rpts)
colnames(rptsdf) <- c("lon", "lat", "occ")
rptsdf$occ <- as.factor(rptsdf$occ)
library(ggplot2)  
p <- ggplot(rptsdf) +
  geom_raster(aes(lon, lat, fill=occ)) +
  scale_fill_manual(values=c("grey95", "darkred", "steelblue"), guide="none") +
  geom_polygon(data = af_adm0, aes(long, lat, group = group), 
               fill = NA, inherit.aes = FALSE) +
  geom_path(data = afss, aes(long, lat, group = group),
            color = "black", inherit.aes = FALSE) +
  coord_equal() + 
  theme_map() + 
  theme(legend.title = element_text(size=12), 
        legend.text = element_text(size=12))

p
ggsave(paste0("figs/occ_grids_", tstamp(), ".png"), p, width=7.7, height=7.3, units="in")
```
